###### Reproducible Research | Peer Assessment 1 
###### Authro: Larry Lugo, Ing. M.Sc.
###### Environment: R x64 v3.1.1. SO: Windows 8.1, 8 MG RAM
###### On an AMD E-300 APU Radeon HD Graphics 1.3 GHz (HP 2000 laptop)
###### Date: October, 2014.

## Summary

This document was created to fulfill Reproducible Research's Peer Assessment 1 requirements on writing a report using **a single R markdown document** that can be processed by **knitr** and be transformed into an HTML file.  

## Setting Working Environment

### Setting global options. Set echo equal to **TRUE** and results equal to **'hold'** for the whole document.  
```{r setoptions}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

### Load required libraries. ggplot was selected for plotting
```{r library_calls}
library(data.table)
library(xtable)
library(ggplot2) 
```

## Getting and Cleaning Data

### Loading and preprocessing the data
According to information provided in this assessment, data is from a personal activity monitoring device that collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual registered during the months of Oct.-Nov., 2012. Additionally, data includes number of steps taken in 5 minute intervals per day.  

Assignment's instructions request to show any code that is needed to loading and preprocessing the data, so the whole process was:

### Load and unzip data

Check if "data" directory (data_dir) exists in working directory (WD). If not the case, create it and download and unzip data info.

```{r file_check}
check_file_exist <- function(file_path) 
{
        if (!file.exists(file_path))
                stop("The ", file_path, " not found!") else TRUE 
}
```

```{r load_data}
load_data <- function(data_dir , fileURL, fileSource) 
{
        # Dataset check and load 
        
        source_path <- paste(data_dir, "/", fileSource , sep="")
        txt_file <- paste(data_dir, "/","activity.csv", sep="")

        if (!file.exists(txt_file)) {
                if (!file.exists(source_path)) {
                        message(paste("Please Wait! Load", fileURL, "..."));
                        download.file(fileURL, destfile=source_path);
                } 
                else {
                    message(paste("Please Wait! Unzip", source_path, " file..."));
                    unzip(source_path, exdir = data_dir);
                }
        }
        message(paste("Please Wait! Load", txt_file, " to dataset..."));
        data <- read.csv(txt_file,
                         header=TRUE,  na.strings="NA",
                         colClasses=c("numeric", "character", "numeric"))
        data$interval <- factor(data$interval)
        data$date <- as.Date(data$date, format="%Y-%m-%d")
        data        
        
}
```

***

### Assign your WD to data_dir variable. Please change it to your own configuration.

```{r Ass_Dir}
data_dir <- "C:/Users/Larry/DataScience-Coursera/ReproducibleResearch/PeerAssessment1";

```

### Load and preparation of *tidy* data

```{r load_prep_call}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" 
fileSource <-"activity.zip"
source_path <- paste(data_dir, "/", fileSource , sep="")
txt_file <- paste(data_dir, "/","activity.csv", sep="")

        if (!file.exists(txt_file)) {
                if (!file.exists(source_path)) {
                        message(paste("Please Wait! Load", fileURL, "..."));
                        download.file(fileURL, destfile=source_path);
                } 
                else {
                    message(paste("Please Wait! Unzip", source_path, " file..."));
                    unzip(source_path, exdir = data_dir);
                }
        }
        message(paste("Please Wait! Load", txt_file, " to dataset..."));
        tidy <- read.csv(txt_file,
                         header=TRUE,   sep=",",
                         colClasses=c("numeric", "character", "numeric"))
        tidy$interval <- factor(tidy$interval)
        tidy$date <- as.Date(tidy$date, format="%Y-%m-%d")
                
```

As a very first and necessary step, see data structure. A preliminary view:  
```{r str_data,}
str(tidy)


## Peer Assessment 1: Questions to be answered

### What is mean total number of steps taken per day?

As instructed by course team, missing values in the data set must be managed to get a tidy data.  

First, a pre-calculation of steps aggregation by day is made:  
```{r pre_calc}
steps_taken_per_day <- aggregate(steps ~ date, tidy, sum)
colnames(steps_taken_per_day) <- c("date", "steps")
```
Then, a histogram of the total number of steps taken each day is plotted with a bin interval of 1000 steps per Day.  
```{r histogram, }
ggplot(steps_taken_per_day, aes(x = steps)) + 
       geom_histogram(fill = "darkblue", binwidth = 1000) + 
        labs(title="Histogram of Steps Taken per Day", 
             x = "Number of Steps per Day", y = "Number of times (Count)") + 
        theme_bw()    
```
  
Finally, mean and median of total number of steps taken per day is showed, removing NA (or missing) values:  

```{r mean_median}
mean_steps = round(mean(steps_taken_per_day$steps, na.rm=TRUE), 2)
median_steps = round(median(steps_taken_per_day$steps, na.rm=TRUE), 2)
```
The Mean is **`r format(mean_steps, digits = 7)`** and 
the Median is **`r format(median_steps,  digits = 7)`**  

Now, next quetion is adressed.

### What is the average daily activity pattern?

Aggregation of steps by intervals of 5-minutes is made, interval to integer is coerced and names are given for colons of the result of this aggregation.  
```{r steps_per_interval}
steps_per_interval <- aggregate(tidy$steps, 
                                by = list(interval = tidy$interval),
                                FUN=mean, na.rm=TRUE)
    # convert to integers for plotting
steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```
Now, a time series of the average number of steps (averaged across all days) versus the 5-minute intervals is plotted:  
```{r plot_steps_per_interval}
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="darkblue", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw() + theme(legend.position = "bottom")
```
  
Next step is find the 5-minute interval containing maximum number of steps:  
```{r max_step_interval}
max_step_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]$interval
```
The variable **`r max_step_interval`<sup>th</sup> 5-minute interval** shows maximum number of steps.  

### Imputing missing values:

#### Total number of missing values in the dataset:
To get this value, proceed as follows: 

```{r TotalNA}
sum(is.na(tidy$steps))
```

#### To get a a new tidy dataset, NA or missing values are replaced by mean values:

```{r fill_data, echo=TRUE}
fill_na <- function(data, defaults) {
        na_indices <- which(is.na(data$steps))
        na_replacements <- unlist(lapply(na_indices, FUN=function(idx){
                interval = data[idx,]$interval
                defaults[defaults$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_indices] <- na_replacements
        fill_steps
}

data_fill <- data.frame(  
        steps = fill_na(tidy, steps_per_interval),  
        date = tidy$date,  
        interval = tidy$interval)
```

#### Plotting a Histogram of number of steps/day  

A bin interval of 1000 steps was chosen:   

```{r complete_steps_per_day, echo=TRUE}

full_steps_per_day <- aggregate(steps ~ date, data_fill, sum)
colnames(full_steps_per_day) <- c("date", "steps")
    
ggplot(full_steps_per_day, aes(x=steps)) + 
        geom_histogram(fill="darkblue", binwidth=1000) + 
        labs(title="Histogram of Full Steps Taken per Day", 
             x="Number of Steps after populate missing values", 
             y="Count") + 
        theme_bw()     

```
  
#### Calculate and report the mean and median total number of steps taken per day  
  
```{r fill_median_mean}
full_mean_steps = round(mean(full_steps_per_day$steps), 2)
full_median_steps = round(median(full_steps_per_day$steps), 2)
```
* Mean after populate missing values is **`r format(full_mean_steps, digits = 7)`**  
* Median populate missing values is **`r format(full_median_steps,  digits = 7)`**

* Mean before populate missing values is **`r format(mean_steps, digits = 7)`**  
* Median before missing values is **`r format(median_steps,  digits = 7)`**

####  What is the impact of imputing missing data on the estimates of the total daily number of steps?

Virtually, median matches the mean.


### Are there differences in activity patterns between weekdays and weekends?

Follow this steps: 

- Add a "day of the week" extra column to the data table.  
- Divide table into two parts: weekends (Saturday and Sunday), and, on the other hand, weekdays (Monday through Friday).  
- Tabulate the average steps per interval for each data set.  
- Plot the two data sets side by side for comparison.  

```{r weekdays}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays in portuguese
    weekend_data <- subset(data, weekday %in% c("sábado","domingo"))
    weekday_data <- subset(data, !weekday %in% c("sábado","domingo"))
    
    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)
    
    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))
    
    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(data_fill)
```

In the next plot, weekdays and weekends steps made by per 5-minute interval are compared: 

```{r plot_weekdays}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="steelblue", size=1) + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()

```
  
## Conclusion:

- A tidy dataset is the first step to get confident and significant results.

- As expected, during weekdays steps are greater than on weekends because of people is moving to their jobs, etc.

- Activity is time-focused on weekends possibly due to free time from jobs.